### 2022-11-30 draft/draft.Rmd
fixed typos in indexing, added more to the abstract

\begin{algorithm}
  \label{algo:pabm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
  \caption{Approximate EM algorithm for the PABM.}
  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
      }
    }
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
        M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
      }
    }
  }
\end{algorithm}

\begin{remark}
Unlike the SBM (when excluding the sparsity factor), each DCBM and PABM is not unique and therefore non-identifiable. 
For example, in the case of the DCBM, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
\citet{Karrer_2011} suggest imposing the constraint $\sum_{i : z_i = k} \omega_i = 1$ for each $k \in \{1, ..., K\}$, but other similar constraints can be used to enforce uniqueness, such as $\max_i \omega_i = 1$ and $\min_i \omega_i = 0$. 
Despite the lack of uniqueness without constraints, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most realizations of the DCBM and PABM, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees. 
Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
\citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results. 
Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters. 
\end{remark}

@@

In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}. 
As in \ref{ex:macaque}, we used $d = 2$ to construct the ASE, then used $K$-curves clustering to fit two curves to these data and predict the labels. 
Figure \ref{fig:macaque-fitted} shows the result of $K$-curves clustering on this graph, 
which yields a 4.4\% error rate.

@@

In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}. 
This graph does not come with ground truth labels. 
\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with $K = 2$ communities. 
Indeed, if we construct an ASE with $p = 1$ and $q = 1$ and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}). 
Nevertheless, we tried fitting an MBM on this graph via $K$-curves clustering. 
Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM. 
\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example. 
The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known *a priori*. 
The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

@@

The RMarkdown and \LaTeX files to compile this dissertation can be found at https://github.com/johneverettkoo/dissertation. 
The code for simulations and data analyses in section \@ref(sec:pabm-grdpg) can be found at https://github.com/johneverettkoo/pabm-grdpg. 
An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at https://github.com/johneverettkoo/osc. 
The code for simulations and data analyses in section \@ref(sec:mbm) can be found at https://github.com/johneverettkoo/manifold-block-models.

>>>>

\begin{algorithm}
  \label{algo:pabm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
  \caption{Approximate EM algorithm for the PABM.}
  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{i \ell} \hat{\lambda}_{j k}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{i \ell} \hat{\lambda}_{j k})) \bigg)$.\;
      }
    }
    \For {$i = 1, ..., n$} {
      \For {$\ell = 1, ..., K$} {
        M-step: $\hat{\lambda}_{i \ell} = \frac{\sum_{j \neq i} \sum_k \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_k \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j k}}$.\;
      }
    }
  }
\end{algorithm}

\begin{remark}
Unlike the SBM (when excluding the sparsity factor), each DCBM and PABM is not unique and therefore non-identifiable. 
For example, in the case of the DCBM, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
\citet{Karrer_2011} suggest imposing the constraint $\sum_{i : z_i = k} \omega_i = 1$ for each $k \in \{1, ..., K\}$, but other similar constraints can be used to enforce uniqueness, such as $\max_i \omega_i = 1$ and $\min_i \omega_i = 0$. 
Despite the lack of uniqueness without constraints, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most realizations of the DCBM and PABM, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees. 
Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
\citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results. 
Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 95\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters. 
\end{remark}

@@

In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}. 
As in example \ref{ex:macaque}, we used $d = 2$ to construct the ASE, then used $K$-curves clustering to fit two curves to these data and predict the labels. 
Figure \ref{fig:macaque-fitted} shows the result of $K$-curves clustering on this graph, 
which yields a 4.4\% error rate.

@@

In the second example, we analyzed the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}. 
Here, the vertices represent the various characters in the *Harry Potter* series, and edges represent whether pairs of characters are enemies. 
This graph does not come with ground truth labels. 
\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with $K = 2$ communities. 
Indeed, if we construct an ASE with $p = 1$ and $q = 1$ and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}). 
Nevertheless, we tried fitting an MBM on this graph via $K$-curves clustering. 
Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM. 
\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example. 
The curves were initialized by specifying that Lord Voldemort (the main antagonist) is in one community and Harry Potter (the titular protagonist) is in another, with the other vertices not known *a priori*. 
The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

@@

The RMarkdown and \LaTeX files to compile this dissertation can be found at https://github.com/johneverettkoo/dissertation. 
The code for simulations and data analyses in section \@ref(sec:pabm-grdpg) can be found at https://github.com/johneverettkoo/pabm-grdpg. 
An R package implementing orthogonal spectral clustering (algorithm \ref{alg:osc}) is currently in progress and can be found at https://github.com/johneverettkoo/osc. 
The code for simulations and data analyses in section \@ref(sec:mbm) can be found at https://github.com/johneverettkoo/manifold-block-models.

### 2022-11-30 draft/draft.tex
fixed typos in indexing, added more to the abstract

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision. One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels). Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then applying clustering methods for these feature vector representations, and finally deriving the asymptotic properties of these methods.

@@

In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}.
As in \ref{ex:macaque}, we used \(d = 2\) to construct the ASE, then used \(K\)-curves clustering to fit two curves to these data and predict the labels.
Figure \ref{fig:macaque-fitted} shows the result of \(K\)-curves clustering on this graph,
which yields a 4.4\% error rate.

@@

In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}.
This graph does not come with ground truth labels.
\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with \(K = 2\) communities.
Indeed, if we construct an ASE with \(p = 1\) and \(q = 1\) and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}).
Nevertheless, we tried fitting an MBM on this graph via \(K\)-curves clustering.
Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM.
\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example.
The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known \emph{a priori}.
The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

@@

The RMarkdown and \LaTeX files to compile this dissertation can be found at \url{https://github.com/johneverettkoo/dissertation}.
The code for simulations and data analyses in section \ref{sec:pabm-grdpg} can be found at \url{https://github.com/johneverettkoo/pabm-grdpg}.
An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at \url{https://github.com/johneverettkoo/osc}.
The code for simulations and data analyses in section \ref{sec:mbm} can be found at \url{https://github.com/johneverettkoo/manifold-block-models}.

>>>>

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from analyzing data in the social sciences to training machine learning models for artificial intelligence tasks. 
One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels). 
Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. 
In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the generalized random dot product graph framework, all block models can be represented as a collection of feature vectors organized by community, applying clustering methods for these feature vector representations by exploiting the linear structures that the block models induce, and finally deriving the asymptotic properties of these methods. 
We further extend this connection between block models and community-organized generalized random dot product graphs to propose more flexible, nonlinear community structures, using real graphs with nonlinear structures as motivating examples.

@@

In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}.
As in example \ref{ex:macaque}, we used \(d = 2\) to construct the ASE, then used \(K\)-curves clustering to fit two curves to these data and predict the labels.
Figure \ref{fig:macaque-fitted} shows the result of \(K\)-curves clustering on this graph,
which yields a 4.4\% error rate.

@@

In the second example, we analyzed the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}.
Here, the vertices represent the various characters in the \emph{Harry Potter} series, and edges represent whether pairs of characters are enemies.
This graph does not come with ground truth labels.
\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with \(K = 2\) communities.
Indeed, if we construct an ASE with \(p = 1\) and \(q = 1\) and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}).
Nevertheless, we tried fitting an MBM on this graph via \(K\)-curves clustering.
Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM.
\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example.
The curves were initialized by specifying that Lord Voldemort (the main antagonist) is in one community and Harry Potter (the titular protagonist) is in another, with the other vertices not known \emph{a priori}.
The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

@@

The RMarkdown and \LaTeX files to compile this dissertation can be found at \url{https://github.com/johneverettkoo/dissertation}.
The code for simulations and data analyses in section \ref{sec:pabm-grdpg} can be found at \url{https://github.com/johneverettkoo/pabm-grdpg}.
An R package implementing orthogonal spectral clustering (algorithm \ref{alg:osc}) is currently in progress and can be found at \url{https://github.com/johneverettkoo/osc}.
The code for simulations and data analyses in section \ref{sec:mbm} can be found at \url{https://github.com/johneverettkoo/manifold-block-models}.
